Microlensing as a possible probe of the internal structure of quasars

### 2015-07-20 paper1/code.tex
small edits
Discussion expanded

Having used to simple model of an ideal fold to gain insight, we now
move to the numerical computation of microlensing light curves with
realistic mass distributions.  For this we we used the microlensing
code by \cite{1999A&A...346L...5W}.

@@

For the computation of the individual deflection angles
$\vec\alpha(\vec\theta)$ a heirarchical or tree method is used.  The
positions of all lensing masses are put into a grid of $\vec\theta$.
Each grid cell is subdivided into four smaller squares recursively
until every cell contains only one mass.  Nearby masses are added
individually, while distance masses are clumped into larger grid cells
whose net contribution is approximated by its first few multipole
moments.  Scaled units are used, with the constant pre-factor in the
deflection angle (\ref{eq:alpha}) separated out.

@@

When generating the magnification map depicted in
Figure~\ref{fig:magnification_map}, which is used in our analysis
below, only two point masses were included.  This was done in order to
have clean fold caustics.  For the computation of the actual
lightcurve, the code was modified to also allow for crescent shaped
images specified through the parameter set $R_p,R_n,a,b$. Here $R_p$
denotes the outer radius of the crescent and $R_n$ the inner one. The
orientation of the source image with respect to the magnification map
is specified by the parameters $a$ and $b$ as the shift of the center
of the inner disk from the center of the outer disk in $x$- and
$y$-direction respectively. In the original version of the code,
gaussian and disk shaped images where already implemented. Those are
completely characterized by the single parameter $R_p$. The values of
the parameters are specified in pixel units corresponding to the
magnification map. Further one needs to specify the start and end
point coordinates of the path, which the center of the source image
follows through the magnification map (see the depiction in figure
\ref{fig:magnification_map}). Hence the points along this path are
specified through the number of timesteps for which the computation of
the brightness is to be carried out. Those points correspond to the
actual measurement of the brightness of an object in the observational
case. For each timestep the two-dimensional convolution of equation
\ref{eqn:ft2d} is carried out numerically for the position of the
source on the magnification map.

>>>>

Having used the simple model of an ideal fold to gain insight, we now
move to the numerical computation of microlensing light curves with
realistic mass distributions.  For this we we used the microlensing
code by \cite{1999A&A...346L...5W}.

@@

For the computation of the individual deflection angles
$\vec\alpha(\vec\theta)$ a heirarchical or tree method is used.  The
positions of all lensing masses are put into a grid of $\vec\theta$.
Each grid cell is subdivided into four smaller squares recursively
until every cell contains only one mass.  Nearby masses are added
individually, while distance masses are clumped into larger grid cells
whose net contribution is approximated by its first few multipole
moments.  Scaled units are used, with the constant pre-factor in the
deflection angle (\ref{eqn:alpha}) separated out.

@@

When generating the magnification map depicted in
Figure~\ref{fig:magnification_map}, which is used in our analysis
below, only two point masses were included.  This was done in order to
have clean fold caustics.  For the computation of the actual
lightcurve, the code was modified to also allow for crescent shaped
images specified through the parameter set $R_p,R_n,a,b$. Here $R_p$
denotes the outer radius of the crescent and $R_n$ the inner one. The
orientation of the source image with respect to the magnification map
is specified by the parameters $a$ and $b$ as the shift of the center
of the inner disk from the center of the outer disk in $x$- and
$y$-direction respectively. In the original version of the code,
gaussian and disk shaped images where already implemented. Those are
completely characterised by the single parameter $R_p$. The values of
the parameters are specified in pixel units corresponding to the
magnification map. Further one needs to specify the start and end
point coordinates of the path, which the center of the source image
follows through the magnification map (see the depiction in figure
\ref{fig:magnification_map}). Hence the points along this path are
specified through the number of timesteps for which the computation of
the brightness is to be carried out. Those points correspond to the
actual measurement of the brightness of an object in the observational
case. For each timestep the two-dimensional convolution of equation
\ref{eqn:ft2d} is carried out numerically for the position of the
source on the magnification map.

### 2015-07-20 paper1/discussion.tex
fitting changes
Discussion expanded

\section{Discussion}
[Mihai's contribution]\\
In the current paper we simulate and study the resulting microlensing lightcurves of geometric crescent-shaped sources 
and compare them with the microlensing lightcurves of other simple mathematically describable source profiles. 
In order to mimic the behaviour of flux of light from the source in the proximity of a fold caustic we make use of the simple approximation described in equation (5). 
The equation would exhaustively describe the magnification map and offer a good universal aproximation for the 
particular microlensing regime that we consider. 
Namely, the shape of the caustic boundary in the proximity of the source in the respective plane can be approximated with
 a line due to reason that the local radius of curvature of caustic is orders of magnitude greater than the half-light
 radius of the studied source. 
In particular cases in which the previously mentioned approximation loses its validity the impact on the quality of the 
lightcurves is not evenly distributed. The shape of the lightcurve will be maintained. 
The datapoints corresponding to the source position before and during the overlapping of the caustic will be affected by
 smaller errors than the datapoints corresponding to later times. \\

The first two source profiles that we consider are the uniform disk and symmetric gaussian source. 
Both of them can be described by a half-light radius $r_{1/2}$ and a total unlensed light flux $S_0$. 
With the two parameters constrained the one dimensional profiles as well as the lightcurves of the two source are completely determined, since no free parameter remains.

@@

Two of the parameters can be reduced by expressing the results in terms of $S_0$ and $r_{1/2}$. 
The later being determinable for any set of parameters $R_p, R_n$ and $a^2+b^2$. Moreover, one of the displacemement 
parameters $b$ has no impact on the one dimensional profile of the source that results from the projection of the source image on an axis perpendicular to the caustic.

Since the one dimensional source profile that corresponds to an axis perpendicular to the caustic contains exhaustively all the information regarding the source that can be revealed by the lightcurve, 
the value of the parameter $b$ does not effect the shape of the lighcurve.

Nevertheless the $b$ parameter is relevant to the calculation of $r_{1/2}$. 
It's qualitative effect is to decrease the value of the half-light radius when the absolute value of the parameter is increased.  
With two parameters constrained and another irrelevant to the shape of the lightcurve, two free parameters remain $R_n$ and $a$. 
Figure 4 reveals that the lighcurve of a crescent source has more visible features than the other two light-curves 
corresponding to the disc and guasssian shape. The parameters $R_n$ and $a$ have strong influences on the shape of 
the microlensed lightcurve, as can be seen in figures 5, 6 and 7. Moreover, the one dimensional source profile corresponding to the direction perpendicular to the caustic reveal four characteristic points. The overlap of each of these points with the caustic leaves visible features on the lightcurve at the corresponding instances of time. In timely order the instances correspond to the start of the overlap between the caustic and the bright disk, the start of the overlap between the caustic and the dark disk, the end of the overlap between the caustic and dark disk and finally the end of the overlap between the caustic and the bright disk.

With the different source profiles and their corresponding lightcurves studied we can change our point of view of the system to that of an observer. The observer would basically detect only the lightcurve of such as source. As described in section (Lightcurve of a crescent source) the timing of the onset and offset of the previously described periods can be used to estimate the values of the radii and one of the displacement parameters when assuming a geometric crescent, the halfwidths and one displacement parameters when assuming a constant brightness elliptical crescent and the halfwidth of the dark ellipse when assuming an elliptical gaussian crescent. All quantities can be estimated in terms of the relative velocity of the source in a direction perpendicular to the caustic.\\

The previously mentioned abstract parameters can be related to physical quantities specific to the central region of a quasar. Thus the luminous region would correspond to the bright accretion disc that surrounds the black hole. The later's gravity would cause a shadow in the bright region limited by the extent of the event horizon of the black hole. Therefore the radius of the bright disk would provide an estimate of the size of the accretion disk and the radius of the dark disc would provide and estimate of the Schwarzschild radius of the black hole. Moreover a value of the Schwarzschild radius 
can be used to estimate the mass of the black hole $M \approx \frac{\Delta t_{dark}  v_p c^2}{2G}$. In the previous 
expression the $\Delta t_{dark}$ denotes the period of overlap between the black hole shadow and fold caustic. 
$v_p$ denotes the component of the relative velocity of the source and fold which is perpendicular to the caustic.  
The respective velocity is an unknown, though it can be constrained on a case by case basis to an order of magnitude or even better. This would require the 
study of the dynamics of the stellar structure which contains the gravitational lens. A better estimate of the relative velocity would facillitate a better estimate of the effective non-rotating black hole mass associated to the black hole shadow.

@@

In case of a high quality lightcurve with insignificant noise and measurement errors, the parameters can be obtained 
by simply identifying the characterstic instances of time withought making use of the actual values of the magnification
 map. If the effect of the errors and the noise distorts the magnification time function enough so that the 
characteristic points are not identifiable with the characteristic periods still visible, the boundaries of the periods 
can be roughly estimated. To study if good estimates of the parameters can still be obtained when datapoints are
 strongly affected by the noise we propose ....

[Irshad's contribution]\\

This work presents a toy model to put useful constraints on the interior structure of quasars, particularly the environment geometry of the event horizon. To carry out this analysis on real data, one needs two situations: First, a quasar which can be observed with a reasonably good telescope with high signal to noise ($\sim 150$) and second, a caustic crossing event of the quasar. These events are not rare, however, to get an isolated caustic crossing might be. As the size of the source is negligible as compared to big caustic structures in the sky, even the single fold caustic crossing is not that rare.

In this toy analysis, we used very ideal data, which might not be the case with real source but we contaminated it to match a reasonable signal to noise.

The main emphasis of this work was the crescent model parameters, which we show can be well recovered by the maximum likelihood analysis, apart from highly degenerate $a$ and $b$ parameters. The reason of this degeneracy is also very intuitive, if one integrates parallel to the caustic to get the integrated 1D brightness profile, so if the motion of the source is either parallel or perpendicular to the caustic, one of the two parameters ($a$ and $b$) is indistinguishable and in case of angled motion of the source, these two are highly degenerate. This degeneracy can be broken if multiple crossing events of the same quasar is available. So for example, one can use the first crossing to estimate the likelihood of the four parameters and use them as priors in the second crossing and so on. In this way, one can accurately and precisely recover all four parameters. Using our toy analysis we conclude that it is feasible, in practice, to recover 2D shape of the source using a microlensing event and distinguish between symmetric and asymmetric sources.

[Contributions from authors]\\
Prasenjit Saha provided the original idea and plan for the research project as well as multiple contributions to the analysis. Mihai Tomozeiu simulated and studied the ideal behaviour of the microlensing lightcurves for the different source profiles discussed. Joachim Wambsganss provided the numerical code used by Manuel Rabold to create the magnification map and corresponding lighcurves used in the MCMC analysis, analisis that was performed by Irshad Mohammed in the last part of the presented work.

>>>>

\section{Discussion}
[Mihai's contribution]\\
In the current paper we simulate and study the resulting microlensing lightcurves of geometric crescent-shaped sources 
and compare them with the microlensing lightcurves of other simple mathematically describable source profiles. 
In order to mimic the behaviour of the flux of light from the source in the proximity of a fold caustic we make use of the simple approximation described in equation (5). 
The equation would exhaustively describe the magnification map and offer a good universal approximation for the 
particular microlensing regime that we consider. 
Namely, the shape of the caustic boundary in the proximity of the source in the respective plane can be approximated with
 a line due to reason that the local radius of curvature of caustic is orders of magnitude greater than the half-light
 radius of the studied source. 
In particular cases in which the previously mentioned approximation loses its validity the impact on the quality of the 
lightcurves is not evenly distributed. The shape of the lightcurve will be maintained. 
The data points corresponding to the source position before and during the overlapping of the caustic will be affected by
 smaller errors than the data points corresponding to later times. \\

The first two source profiles that we consider are the uniform disk and symmetric Gaussian source. 
Both of them can be described by a half-light radius $r_{1/2}$ and a total unlensed light flux $S_0$. 
With the two parameters constrained the one dimensional profiles as well as the lightcurves of the two source are completely determined, since no free parameter remains.

@@

Two of the parameters can be reduced by expressing the results in terms of $S_0$ and $r_{1/2}$. 
The later being determinable for any set of parameters $R_p, R_n$ and $a^2+b^2$. Moreover, one of the displacement 
parameters $b$ has no impact on the one dimensional profile of the source that results from the projection of the source image on an axis perpendicular to the caustic.

Since the one dimensional source profile that corresponds to an axis perpendicular to the caustic contains exhaustively all the information regarding the source that can be revealed by the lightcurve, 
the value of the parameter $b$ does not have an effect on the shape of the lighcurve.

Nevertheless the $b$ parameter is relevant for the calculation of $r_{1/2}$. 
It's qualitative effect is to decrease the value of the half-light radius when the absolute value of the parameter is increased.  
With two parameters constrained and another irrelevant to the shape of the lightcurve, two free parameters remain $R_n$ and $a$. 
Figure 4 reveals that the lighcurve of a crescent source has more visible features than the other two light-curves 
corresponding to the disc and Guasssian shape. The parameters $R_n$ and $a$ have strong influences on the shape of 
the microlensed lightcurve, as can be seen in figures 5, 6 and 7. Moreover, the one dimensional source profile corresponding to the direction perpendicular to the caustic reveal four characteristic points. The overlap of each of these points with the caustic leaves visible features on the lightcurve at the corresponding instances of time. In timely order the instances correspond to the start of the overlap between the caustic and the bright disk, the start of the overlap between the caustic and the dark disk, the end of the overlap between the caustic and dark disk and finally the end of the overlap between the caustic and the bright disk.

With the different source profiles and their corresponding lightcurves studied we can change our point of view of the system to that of an observer. The observer would basically detect only the lightcurve of such as source. As described in section 4.3 (Lightcurve of a crescent shaped source) the timing of the onset and offset of the previously described periods can be used to estimate the values of the radii and one of the displacement parameters when assuming a geometric crescent. All quantities can be estimated in terms of the relative velocity of the source in a direction perpendicular to the caustic.\\

The previously mentioned abstract parameters can be related to physical quantities specific to the central region of a quasar. Thus the luminous region would correspond to the bright accretion disc that surrounds the black hole. The later's gravity would cause a shadow in the bright region limited by the extent of the event horizon of the black hole. Therefore the radius of the bright disk would provide an estimate of the size of the accretion disk and the radius of the dark disc would provide and estimate of the Schwarzschild radius of the black hole. Moreover a value of the Schwarzschild radius 
can be used to estimate the mass of the black hole $M \approx \frac{\Delta t_{dark}  v_p c^2}{2G}$. In the previous 
expression the $\Delta t_{dark}$ denotes the period of overlap between the black hole shadow and fold caustic. 
$v_p$ denotes the component of the relative velocity of the source and fold which is perpendicular to the caustic.  
The respective velocity is an unknown, though it can be constrained on a case by case basis to an order of magnitude or even better. This would require the 
study of the dynamics of the stellar structure which contains the gravitational lens. A better estimate of the relative velocity would facilitate a better estimate of the effective non-rotating black hole mass associated to the black hole shadow.

@@

In case of a high quality lightcurve with insignificant noise and measurement errors, the parameters can be obtained 
by simply identifying the characteristic instances of time without making use of the actual values of the magnification
 map. If the effect of the errors and the noise distorts the magnification time function enough so that the 
characteristic epochs are not identifiable with the characteristic periods still visible, the boundaries of the periods 
can be roughly estimated. Furthermore, if direct estimates of the parameters cannot be obtained we propose the use of a strong statistical tool such as Markov-Chain Monte Carlo. In Section 6  we have studied the possibility of identifying a crescent source and  the possibility of recovering the respective parameters. As magnification map we have used a complex numerical one generated with the microlensing code by \cite{1999A&A...346L...5W}. In addition we have added to the signal a Gaussian  noise with a SNR of 1.6.   In the experiment we have considered all nine combinations of original source profiles and assumed fitting source models. Effectively we have fitted using MCMC all three sources with all three fitting assumed models. The results of the experiment allowed us build a procedure for distinguishing the shape of the source assuming the one of the three models we have considered are a good approximation.  The procedure would be useful for observers that endeavour to gain more information about an unresolved source for which they can study the microlensing lightcurve.  As a first step in the procedure, one should first attempt to use MCMC with a Gaussian model assumption. If either the value of the $\chi^2$ or the number of rejected datapoints is large then the next step is to change the assumed source model to an uniform disc and redo the fitting. Finally if the uniform disc assumption is rejected as well then the fitting should be done with a crescent source model assumption. If at each of the three steps the data rejects the model than the source cannot be approximated by none of the three models. Otherwise if at one of the steps the data does not reject the model then the respective model is a good approximation.

[Contributions from authors]\\
Prasenjit Saha provided the original idea and plan for the research project as well as multiple contributions to the analysis. Mihai Tomozeiu simulated and studied the ideal behaviour of the microlensing lightcurves for the different source profiles discussed. Joachim Wambsganss provided the numerical code used by Manuel Rabold to create the magnification map and corresponding lighcurves used in the MCMC analysis, analysis that was performed by Irshad Mohammed in the last part of the presented work.
